Penalized linear mixed models for structured genetic data

### 2021-03-12 main.tex
Edits to binary paragraph, lasso section, and software

\anna{In many genetic association studies the phenotype of interest is binary, for example the presence or absence of a heritable trait. However, the extension of LMMs and penalized LMMs to nonlinear models is not straightforward. LMMs and penalized LMMs are able to efficiently correct for dependencies among subjects by preprocessing the data by means of a weighted rotation. This procedure does not generalize beyond the linear regression case where the outcome is continuous. Because of this, distinct inferential algorithms are required to extend the LMM modeling methodology to nonlinear models such as logistic regression. \citet{mandt2017sparse} generalize the LMM modeling paradigm to binary phenotypes in the interest of classification. They do so by thresholding the outcome variable through a Probit likelihood with additional sparsity constraints and approximate Bayesian inference. However, to the best of our knowledge, the joint problem of sparse estimation in the presence of dependent observations has been limited to the linear regression case \citep{seeger2011large, vattikuti2014applying, rakitsch2013lasso}.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods} \label{sec:methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
\subsection{\anna{lasso}}
\anna{not sure if you/reviewers were thinking more lasso background should go here vs. somewhere else. I initially tried to add more background to the introduction but didn't think additional information "fit" there given what was already there. I'm open to having additional lasso background elsewhere though.}
We begin by briefly reviewing the key methodological components of the lasso. In the usual linear regression set-up where there is a continuous response $\by \in \real^n$, an $n \times p$ design matrix $\bX$, and a parameter vector $\bbeta \in \real^p$, the ordinary lasso minimizes the objective function

\begin{equation}
    Q(\bbeta | \lam) = \frac{1}{2n}|| \by -  \beta_0 - \bX \bbeta ||_2^2 + \lam ||\bbeta||_1,
\end{equation}
where $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients, and $||\mathbf{x}||_2 = \sqrt{\sum_{i=1}^p x_i^2}$ the $\ell_2$ norm. The first component of the objective function is the linear regression loss function, which quantifies the overall model fit and captures how well the estimated parameters agree with the observed data. As in ordinary least squares (OLS) regression, the lasso assumes observations are independent. The second component is the $\ell_1$ penalty applied to the estimated coefficient values (but not the intercept, $\beta_0$), which discourages large, unrealistic parameter values. The tuning parameter, $\lambda$, controls the tradeoff between these two components. For sufficiently large values of $\lambda$ all elements of $\bbeta$ will be estimated to be 0. In the $p \ll n$ setting, if $\lam = 0$, the lasso solution coincides with that of OLS. The $\ell_1$ penalty of the lasso can be applied to nonlinear models such as Cox regression \citep{tibshirani1997lasso} and logistic regression \citep{shevade2003simple} by replacing the residual sum of squares with the appropriate negative log-likelihood function.

@@

Different variance component estimation methods lead to different estimates of $\bW$ and two distinct LMM-lasso variants: LMM-lasso-Rakitsch and LMM-lasso-ggmix. LMM-lasso-Rakitsch uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmass$ \citep{rakitsch2013lasso}. These variance components and the corresponding weights are estimated once under the assumption of a null model. A lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \citep{glmnet} or \texttt{ncvreg} \citep{ncvreg}. LMM-lasso-ggmix iteratively updates $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm \citep{bhatnagar2020simultaneous}. This iterative update scheme prevents the use of standard lasso software, but is implemented in the \texttt{R} package \texttt{ggmix}. \anna{Should we still reference glmnet, ncvreg, ggmix here if we now include a software section?} We further discuss differences between these methods in Section \ref{sec:results}.

@@

>>>>

\pb{Please read the following paragraph; it has been modified extensively.}

In many genetic association studies the phenotype of interest is binary, for example the presence or absence of a heritable trait. Unfortunately, the extension of LMMs and penalized LMMs to nonlinear models is not straightforward. As we explain in Section~\ref{sec:methods}, LMMs and penalized LMMs are able to efficiently correct for dependencies among subjects by preprocessing the data by means of a weighted rotation. This procedure does not directly generalize beyond the linear regression case. The situation is certainly not impossible, and many approaches to fitting generalized linear mixed models, or GLMMs, have been proposed in the classical, low-dimensional setting. However, very little work has been done to extend this to penalized, high-dimensional models and to the best of our knowledge there is no existing software for doing so at this time. We hope that this situation changes in the near future.

\pb{What is the significance of this Mandt paper? Isn't there a huge literature on GLMMs? \citet{mandt2017sparse} generalize the LMM modeling paradigm to binary phenotypes in the interest of classification. They do so by thresholding the outcome variable through a Probit likelihood with additional sparsity constraints and approximate Bayesian inference. }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods} \label{sec:methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
\subsection{Lasso}
We begin by briefly reviewing the key methodological components of the lasso. In the usual linear regression setting involving a continuous response $\by \in \real^n$, an $n \times p$ design matrix $\bX$, and a parameter vector $\bbeta \in \real^p$, the ordinary lasso minimizes the objective function

\begin{equation}
    Q(\bbeta | \lam) = \frac{1}{2n}|| \by -  \beta_0 - \bX \bbeta ||_2^2 + \lam ||\bbeta||_1,
\end{equation}
where $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients, and $||\mathbf{x}||_2 = \sqrt{\sum_{i=1}^p x_i^2}$ the $\ell_2$ norm. The first component of the objective function is the linear regression loss function, which quantifies the overall model fit and captures how well the estimated parameters agree with the observed data (here, assuming that observations are independent). The second component is the $\ell_1$ penalty applied to the estimated coefficient values (but not the intercept, $\beta_0$), which discourages large, unrealistic parameter values. The tuning parameter, $\lambda$, controls the tradeoff between these two components. For sufficiently large values of $\lambda$ all elements of $\bbeta$ will be estimated to be 0. At the other extreme, if $\lam = 0$ the lasso solution coincides with that of classical ordinary least squares regression. The $\ell_1$ penalty of the lasso can be applied to nonlinear models such as Cox regression \citep{tibshirani1997lasso} and logistic regression \citep{shevade2003simple} by replacing the residual sum of squares with the appropriate loss function (i.e., the negative log-likelihood).

The lasso penalty accomplishes two key things: (1) shrinkage, meaning that coefficients are encouraged to be smaller and less likely to be overestimated and (2) sparsity, meaning that many coefficients will be exactly zero and thus eliminated from the model. Both of these components are key to the lasso's ability to work with large, genome-scale data. The following sections discuss how to lasso can be modified to account for correlation among individuals.

\pb{additional references?  Elements of statistical learning?}

@@

Different variance component estimation methods lead to different estimates of $\bW$ and two distinct LMM-lasso variants: LMM-lasso-Rakitsch and LMM-lasso-ggmix. LMM-lasso-Rakitsch uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmass$ \citep{rakitsch2013lasso}. These variance components and the corresponding weights are estimated once under the assumption of a null model. The resulting optimization problem is thus reduced to that of the ordinary lasso and any software can be used to estimate $\bbeta$. LMM-lasso-ggmix iteratively updates $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm \citep{bhatnagar2020simultaneous}, and thus requires special software. Section~\ref{sec:software} discusses software packages for fitting these models in greater detail.

@@

\pb{BELOW IS JUST DUMP FROM EARLIER}
. This iterative update scheme prevents the use of standard lasso software, but is implemented in the \texttt{R} package

A lasso model can then be fit using standard software for

A lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \citep{glmnet} or \texttt{ncvreg} \citep{ncvreg}.

LMM-lasso-ggmix iteratively updates $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm \citep{bhatnagar2020simultaneous}. This iterative update scheme prevents the use of standard lasso software, but is implemented in the \texttt{R} package \texttt{ggmix}. We further discuss differences between these methods in Section \ref{sec:results}.
\pb{END DUMP}
